---
title: "Introduction to WOVEN: Multi-Omics Integration for Incomplete Patient Data"
author: "Nathan Bresette, Ai-Ling Lin, Jianlin Cheng"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Introduction to WOVEN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Overview

**WOVEN** (Weighted Omics View Embedding via Nystrom) is a supervised
multi-omics integration method designed for clinical cohorts where patients
are missing entire assay blocks. Standard methods such as `mixOmics::DIABLO`
enforce a strict intersection constraint: every patient must have every
modality, so patients missing even one block are discarded. In a typical
three-platform study this can eliminate 50--80% of enrolled subjects,
systematically excluding the sickest patients who are most likely to be
missing a blood draw or imaging scan.

WOVEN solves this by:

1. Learning shared projection matrices $W_v$ using only **anchor subjects**
   (fully-observed complete cases), solving a supervised SUMCOR MCCA in dual
   space via a single closed-form eigendecomposition.
2. Projecting **block-missing subjects** into the shared latent space using
   their available modalities directly ($z_i = x_i W_v$), or via Nyström
   extension when an entire modality block is absent.

No feature-level imputation is performed at any stage.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("woven")
```

# Quick Start

## Simulate a small three-modality dataset

```{r simulate}
library(woven)
set.seed(42)

n  <- 120   # total subjects
p1 <- 200   # modality 1 features (e.g. RNA-seq)
p2 <- 80    # modality 2 features (e.g. methylation)
p3 <- 30    # modality 3 features (e.g. proteomics)
K  <- 3     # latent dimensions
n_groups <- 3

# True group labels
labels <- rep(1:n_groups, each = n / n_groups)

# Simulate group-discriminative signal in each modality
make_view <- function(p, labels, signal = 2) {
  n <- length(labels)
  X <- matrix(rnorm(n * p), n, p)
  for (g in unique(labels))
    X[labels == g, 1:5] <- X[labels == g, 1:5] + signal
  X
}

X1 <- make_view(p1, labels)
X2 <- make_view(p2, labels, signal = 3)
X3 <- make_view(p3, labels, signal = 4)

X_list <- list(rna = X1, methyl = X2, protein = X3)
```

## Induce block-level missingness

```{r missingness}
# 40% MCAR block missingness -- every patient retains >= 1 modality
set.seed(7)
V <- length(X_list)
miss_mask <- matrix(runif(n * V) < 0.4, n, V)

# Enforce >= 1 view per patient
all_miss <- which(rowSums(miss_mask) == V)
for (i in all_miss) miss_mask[i, sample(V, 1)] <- FALSE

# Anchor subjects: fully observed
anchor_idx <- which(rowSums(miss_mask) == 0L)
cat(sprintf("Anchors: %d / %d (%.0f%%)\n",
    length(anchor_idx), n, 100 * length(anchor_idx) / n))

# Apply missingness
X_miss <- lapply(seq_len(V), function(v) {
  Xv <- X_list[[v]]
  Xv[miss_mask[, v], ] <- NA_real_
  Xv
})
names(X_miss) <- names(X_list)
```

## Fit WOVEN

```{r fit}
fit <- woven(
  X_list     = X_miss,
  Y          = labels,
  anchor_idx = anchor_idx,
  K          = K,
  lambdas    = 0.01,   # graph Laplacian regularization
  gamma_y    = 5.0,    # label supervision strength
  k_nn       = 10L     # k-NN graph for Laplacian
)

fit  # print summary
```

## Inspect the latent space

`fit$Z` is an n x K matrix of consensus latent scores for **all** subjects,
including the block-missing ones. No subject is dropped.

```{r plot, fig.cap="WOVEN latent space. Filled points are anchor subjects (complete cases); open circles are block-missing subjects projected from their available views. Colors indicate true group membership."}
plot(fit, labels = labels,
     main = sprintf("WOVEN latent space — %d/%d subjects scored", n, n))
```

## Score new subjects

`woven_scores()` projects a new cohort into the same latent space and
returns an n_new x K matrix directly.

```{r scores_new}
# Simulate 20 new subjects with random block missingness
set.seed(99)
X_new <- lapply(X_list, function(Xv) Xv[1:20, ])
miss_new <- matrix(runif(20 * V) < 0.4, 20, V)
for (i in 1:20) if (all(miss_new[i,])) miss_new[i, sample(V,1)] <- FALSE
for (v in seq_len(V)) X_new[[v]][miss_new[, v], ] <- NA_real_

Z_new <- woven_scores(fit, X_new)
dim(Z_new)   # 20 x K
```

## Predict classes for new subjects

`woven_predict()` returns class probabilities alongside latent scores.

```{r predict}
pred <- woven_predict(fit, X_new)
head(pred[, 1:3])
```

# Choosing Hyperparameters

WOVEN has three hyperparameters. Defaults work well in practice (confirmed
by sensitivity analysis across four orders of magnitude in the paper):

| Parameter | Default | Role |
|-----------|---------|------|
| `lambdas` | 0.1     | Graph Laplacian regularization strength |
| `gamma_y` | 1.0     | Label supervision strength (0 = unsupervised CCA) |
| `k_nn`    | 15      | k-NN graph connectivity for Laplacian |

Setting `gamma_y = 0` reduces WOVEN to unsupervised dual CCA -- the
supervision term is essential for discriminative performance.

# Comparison with DIABLO

The key advantage of WOVEN over `mixOmics::DIABLO` is effective sample
size (ESS): DIABLO discards every subject missing any modality, while WOVEN
retains all subjects with at least one observed view.

```{r ess}
n_diablo <- length(anchor_idx)       # DIABLO can only score complete cases
n_woven  <- sum(!is.na(fit$Z[,1]))  # WOVEN scores all subjects with >= 1 view

cat(sprintf("DIABLO ESS: %d / %d (%.0f%%)\n", n_diablo, n, 100*n_diablo/n))
cat(sprintf("WOVEN  ESS: %d / %d (%.0f%%)\n", n_woven,  n, 100*n_woven/n))
```

# Session Info

```{r session}
sessionInfo()
```
