---
title: "Introduction to WOVEN: Multi-Omics Integration for Incomplete Patient Data"
author: "Nathan Bresette, Ai-Ling Lin, Jianlin Cheng"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to WOVEN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment  = "#>",
    fig.width  = 8,
    fig.height = 5,
    warning  = FALSE,
    message  = FALSE
)
```

# Overview

**WOVEN** (Weighted Omics View Embedding via Nystrom) is a supervised
multi-omics integration method designed for clinical cohorts where patients
are missing entire assay blocks. Standard methods such as `mixOmics::DIABLO`
enforce a strict intersection constraint: every patient must have every
modality, so patients missing even one block are discarded. In a typical
three-platform study this can eliminate 50--80% of enrolled subjects,
systematically excluding the sickest patients.

WOVEN solves this by learning projection matrices W from **anchor subjects**
(fully-observed complete cases), then projecting block-missing subjects using
their available views. No feature-level imputation is performed.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("woven")
```

# Preparing Your Data

## Input format

`woven()` takes a named list of matrices --- one per omics platform --- where:

- Rows are **subjects** (same order across all matrices)
- Columns are **features** (genes, CpGs, proteins, metabolites, etc.)
- Subjects missing an **entire modality block** have that row set to `NA`

```{r data-format, eval=FALSE}
# Correct format: row = subject, col = feature
# Subjects missing a platform have that entire row set to NA
X_list <- list(
    RNA   = rna_matrix,    # n x p_rna,  missing rows = all NA
    Methyl = methyl_matrix, # n x p_meth, missing rows = all NA
    Prot  = prot_matrix    # n x p_prot, missing rows = all NA
)
```

## Pre-processing recommendations

**WOVEN expects pre-processed, approximately continuous data.** Apply standard
platform-specific normalization before passing data to `woven()`:

| Platform | Recommended pre-processing |
|---|---|
| RNA-seq (bulk) | `log1p(CPM)` then `scale()` per gene |
| DNA methylation | M-values (`log2(beta/(1-beta))`), then `scale()` |
| Proteomics (LC-MS) | `log2(intensity)`, median-center per sample |
| Metabolomics | `log2` transform, then `scale()` |
| Microbiome (16S) | CLR transform (e.g. `compositions::clr()`) |

Raw counts, raw beta values, or raw intensities will produce poorly scaled
projection matrices. When in doubt, `scale()` each modality matrix so all
features have mean 0 and SD 1.

```{r preprocess, eval=FALSE}
# Example: log1p + scale for RNA-seq
X_rna_scaled <- scale(log1p(raw_counts))

# Then set block-missing rows to NA (do NOT impute)
X_rna_scaled[subjects_missing_rna, ] <- NA
```

# Quick Start

## Simulate a three-modality dataset

```{r simulate}
library(woven)
set.seed(42)

n       <- 150   # total subjects
K       <- 3     # latent dimensions to learn
n_groups <- 3

# True group labels (e.g. CN, MCI, Dementia)
Y <- factor(rep(c("CN", "MCI", "Dementia"), each = n / n_groups),
            levels = c("CN", "MCI", "Dementia"))

# Simulate three pre-processed modality matrices with group signal
make_modality <- function(p, Y, signal = 3) {
    X <- matrix(rnorm(length(Y) * p), length(Y), p)
    for (g in levels(Y))
        X[Y == g, seq_len(8)] <- X[Y == g, seq_len(8)] + signal
    colnames(X) <- paste0("Feature_", seq_len(p))
    X
}

X_rna   <- make_modality(300, Y, signal = 3)
X_meth  <- make_modality(120, Y, signal = 4)
X_prot  <- make_modality(40,  Y, signal = 5)
```

## Induce block-level missingness

```{r missingness}
set.seed(7)
V        <- 3
miss_mask <- matrix(runif(n * V) < 0.35, n, V)

# Every subject must retain at least one modality
for (i in which(rowSums(miss_mask) == V))
    miss_mask[i, sample(V, 1)] <- FALSE

# Apply missingness: entire row -> NA for missing modalities
X_rna[ miss_mask[, 1], ] <- NA
X_meth[miss_mask[, 2], ] <- NA
X_prot[miss_mask[, 3], ] <- NA

n_anchors <- sum(rowSums(miss_mask) == 0)
cat(sprintf(
    "Anchors (all 3 views): %d / %d (%.0f%%)\n",
    n_anchors, n, 100 * n_anchors / n
))
```

## Fit WOVEN

Pass a **named** list so modality labels appear in all plots automatically.
`anchor_idx` is optional -- WOVEN detects anchors from the NA pattern.

```{r fit}
fit <- woven(
    X_list  = list(RNA = X_rna, Methylation = X_meth, Proteomics = X_prot),
    Y       = Y,
    K       = K,
    lambdas = 0.01,   # graph Laplacian regularization (robust to this)
    gamma_y = 5.0,    # label supervision strength (any value > 0 works)
    k_nn    = 10L     # k-NN graph for Laplacian
)

fit   # prints summary + guided next steps
```

# Exploring Results

## Latent space scatter

`plot()` shows all scored subjects colored by group. Solid points are anchors;
faded points are block-missing subjects projected from their available views.
Ellipses show the 68% confidence region per group (anchors only).

```{r plot-scatter, fig.cap="WOVEN latent space. All subjects are scored regardless of missingness."}
plot(fit, labels = Y)
```

## VIP scores: which features matter most?

Variable Importance in Projection (VIP) scores rank features by their
contribution to the shared latent space across all K dimensions. VIP > 1
indicates above-average importance.

```{r plot-vip, fig.cap="Top RNA features by VIP score."}
woven_plot_vip(fit, modality = "RNA", n_top = 15)
```

Plot VIP for a different modality:

```{r plot-vip2, fig.cap="Top methylation features by VIP score."}
woven_plot_vip(fit, modality = "Methylation", n_top = 15)
```

## Feature loadings: direction of effect

Loadings show which features push subjects in the positive vs. negative
direction along a given latent dimension. Equivalent to `plotLoadings()` in
DIABLO.

```{r plot-loadings, fig.width=10, fig.cap="Top feature loadings for latent dimension 1 across all three modalities."}
woven_plot_loadings(fit, dim = 1L, n_top = 10)
```

## Variance explained: choosing K

Use this to verify that your chosen K captures most of the shared signal.
If the last dimension still explains substantial variance, increase K.

```{r plot-variance, fig.cap="Proportion of shared variance explained per latent dimension."}
woven_plot_variance(fit)
```

## Quantitative metrics

`woven_metrics()` computes silhouette score, Davies-Bouldin index, NMI, and
effective sample size (ESS) retention directly from the fit object.

```{r metrics}
woven_metrics(fit, Y)
```

ESS = 1.00 means every subject with at least one observed modality was scored.
Compare this to DIABLO, which can only score subjects present in all modalities.

# Predicting New Subjects

```{r predict}
# Simulate 20 new subjects with partial missingness
set.seed(99)
X_new <- list(
    RNA         = make_modality(300, Y[1:20], signal = 3),
    Methylation = make_modality(120, Y[1:20], signal = 4),
    Proteomics  = make_modality(40,  Y[1:20], signal = 5)
)
# Some new subjects are missing a modality
X_new$RNA[1:5, ]        <- NA
X_new$Proteomics[6:10, ] <- NA

pred <- woven_predict(fit, X_new)
head(pred[, 1:3])
```

`predicted_class` returns original label names ("CN", "MCI", "Dementia"),
not integer codes.

# Comparing WOVEN and DIABLO: Effective Sample Size

The key advantage of WOVEN over `mixOmics::DIABLO` is **effective sample size
(ESS)**. DIABLO requires every subject to have every modality; WOVEN scores
all subjects with at least one observed view.

```{r ess-comparison}
n_anchors <- length(fit$anchor_idx)    # what DIABLO would use
n_woven   <- sum(!is.na(fit$Z[, 1]))  # what WOVEN scores

cat(sprintf("DIABLO ESS: %d / %d (%.0f%%)\n",
            n_anchors, n, round(100 * n_anchors / n)))
cat(sprintf("WOVEN  ESS: %d / %d (%.0f%%)\n",
            n_woven,   n, round(100 * n_woven   / n)))
```

In real cohorts, the subjects DIABLO discards are often the sickest patients
(those who miss a blood draw or imaging scan due to disease severity),
introducing systematic bias in comparative effectiveness estimates.

# Accessing Raw Results

The fit object exposes all internal quantities for custom downstream analyses:

```{r access}
dim(fit$Z)          # n x K consensus latent scores for ALL subjects
names(fit$W_list)   # one projection matrix per modality
dim(fit$W_list$RNA) # p_rna x K

# Row names on Z come from row names of your input matrices
# (set rownames on X_list matrices to get named rows in fit$Z)
```

`fit$Z` can be passed directly to survival analysis, clustering, or any
downstream model as a low-dimensional representation of the full cohort.

# Session Info

```{r session}
sessionInfo()
```
